Notes

### 2023-11-21 chapters/background-related-work.tex
Update reference

Turning our attention to logistic regression, a similar \(L^2\)-based functional logistic equation can be derived for the binary classification problem via the logistic function:
\begin{equation}\label{eq:l2-logistic-model}
  \P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\beta}\}},
\end{equation}
where \(\alpha_0 \in \R\) and \(\beta \in L^2[0, 1]\). In this situation, the most common way of estimating the slope function \(\beta\) is via its Maximum Likelihood Estimator (MLE). However, not only do the same complications as in the linear regression model apply in this situation, but there is also the additional problem that in functional settings the MLE does not exist with probability one under fairly general conditions \citep[see][]{berrendero2018functional}.

@@

In spite of the apparent generality of model~\eqref{eq:l2-linear-model}, it can be shown that it is not flexible enough to include ``simple'' finite-dimensional models based on linear combinations of the marginals of the process, such as \(Y=\alpha_0 + \beta_1 X(t_1)+ \cdots + \beta_p X(t_p) + \epsilon\) for some constants \(\beta_j\in\R\) and instants \(t_j\in[0,1]\); see \citet{berrendero2020general} for additional details on this. It turns out that in both linear and logistic scenarios a natural and more general alternative to the \(L^2\)-model is the so-called Reproducing Kernel Hilbert Space (RKHS) model, which instead assumes the unknown functional parameter to be a member of the RKHS associated with the covariance function of the process \(X\), making use of the scalar product of that space. As we will show later on, not only is this model simpler and arguably easier to interpret, but it also constrains the parameter space to smoother and more manageable functions. In fact, it does include a model based on finite linear combinations of the marginals of \(X\) as a particular case, which is especially appealing to practitioners confronted with functional data problems due to its simplicity. These RKHS-based models and their idiosyncrasies have been explored in \citet{berrendero2019rkhs, berrendero2020general} in the functional linear regression setting, and in \citet{berrendero2018functional} for the case of functional logistic regression.

>>>>

Turning our attention to logistic regression, a similar \(L^2\)-based functional logistic equation can be derived for the binary classification problem via the logistic function:
\begin{equation}\label{eq:l2-logistic-model}
  \P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\beta}\}},
\end{equation}
where \(\alpha_0 \in \R\) and \(\beta \in L^2[0, 1]\). In this situation, the most common way of estimating the slope function \(\beta\) is via its Maximum Likelihood Estimator (MLE). However, not only do the same complications as in the linear regression model apply in this situation, but there is also the additional problem that in functional settings the MLE does not exist with probability one under fairly general conditions \citep[see][]{berrendero2023functional}.

@@

In spite of the apparent generality of model~\eqref{eq:l2-linear-model}, it can be shown that it is not flexible enough to include ``simple'' finite-dimensional models based on linear combinations of the marginals of the process, such as \(Y=\alpha_0 + \beta_1 X(t_1)+ \cdots + \beta_p X(t_p) + \epsilon\) for some constants \(\beta_j\in\R\) and instants \(t_j\in[0,1]\); see \citet{berrendero2020general} for additional details on this. It turns out that in both linear and logistic scenarios a natural and more general alternative to the \(L^2\)-model is the so-called Reproducing Kernel Hilbert Space (RKHS) model, which instead assumes the unknown functional parameter to be a member of the RKHS associated with the covariance function of the process \(X\), making use of the scalar product of that space. As we will show later on, not only is this model simpler and arguably easier to interpret, but it also constrains the parameter space to smoother and more manageable functions. In fact, it does include a model based on finite linear combinations of the marginals of \(X\) as a particular case, which is especially appealing to practitioners confronted with functional data problems due to its simplicity. These RKHS-based models and their idiosyncrasies have been explored in \citet{berrendero2019rkhs, berrendero2020general} in the functional linear regression setting, and in \citet{berrendero2023functional} for the case of functional logistic regression.

### 2023-11-21 chapters/bayesian-functional-regression.tex
Update reference

In this chapter we present the precise functional models and Bayesian methodologies explored in this work. The RKHS-based functional models under consideration \citep[see][]{berrendero2018functional, berrendero2019rkhs} are those obtained by substituting the functional parameter \(\beta\in L^2[0,1]\) for \(\alpha \in \Hcal(K)\), replacing also the scalar product \(\dotprod{X}{\beta}\) for \(\dotprod{X}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subspace \(\Hcal_0(K)\) defined in~\eqref{eq:h0}, i.e.:
\begin{equation}\label{eq:alpha_h0}
\alpha(\cdot) = \sum_{j=1}^p \beta_j K(t_j, \cdot), \text{ for some } p \in \N, \ \beta_j \in \R \text{ and } t_j \in [0,1].
\end{equation}
Moreover, as we said before, with a slight abuse of notation we will understand the expression \(\dotprod{x}{\alpha}_K\) as \(\Psi_x(\alpha)\), where \(x=X(\omega)\) and \(\Psi_x\) is Loève's isometry. Hence, taking into account that \(\Psi_X(K(t, \cdot)) = X(t)\) by definition (see~\eqref{eq:loeves-isometry}), when \(\alpha\) is as in~\eqref{eq:alpha_h0} we can write \(\dotprod{x}{\alpha}_K \equiv \sum_{j=1}^p \beta_j x(t_j)\).

In this way we get a simpler, finite-dimensional approximation of the functional RKHS model, which we argue reduces the overall complexity of the model while still capturing most of the relevant information. When it comes to parameter estimation, a direct optimization of some loss function would probably require a tailored algorithm that took into account the whole functional trajectories \(x(t)\) to select the appropriate times \(t_j\). Indeed, such an idea is explored in \citet{berrendero2018functional} for the logistic regression case, where the authors propose a ``greedy max-max'' method reminiscent of the EM algorithm that alternates between estimating the coefficients and the time instants through a maximum likelihood approach.

@@

In the case of functional logistic regression, we regard the response variable \(Y\in\{0, 1\}\) as a Bernoulli random variable given \(X=x \in L^2[0, 1]\), and as usual suppose that \(\log\left(p(x)/(1-p(x))\right)\) is linear in \(x\), where \(p(x)=\P(Y=1| X=x)\). Then, following the approach  suggested by \citet{berrendero2018functional}, a RKHS model might be given, in terms of the correspondence \(\dotprod{X}{\alpha}_K = \Psi_X(\alpha)\), by the equation
\begin{equation}\label{eq:rkhs-model-logistic}
  \P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\alpha}_K\}}, \quad \alpha_0 \in \R, \ \alpha \in \Hcal_{0,p}(K).
\end{equation}

@@

In much the same way as the linear regression model described above, this RKHS-based logistic regression model offers some advantages over the \(L^2\)-model~\eqref{eq:l2-logistic-model}. First and foremost, it has a more straightforward interpretation and allows for a workable Bayesian approach, as we will demonstrate below. Secondly, it can be shown that under mild conditions the general RKHS functional logistic model holds whenever the conditional distributions \(X | Y=0\) and \(X|Y=1\) are homoscedastic Gaussian processes, and in some cases it also entails the \(L^2\)-model \citep[see Theorem 1 in][]{berrendero2018functional}; this provides a sound theoretical motivation for the reduced model. Furthermore, a maximum likelihood approach for parameter estimation (although not considered here) is possible as well. Indeed, the use of a finite-dimensional approximation  mitigates the problem of non-existence of the MLE in the functional case. However, let us recall that even in finite-dimensional settings there are cases of quasi-complete separation in which the MLE does not exist \citep{albert1984existence}, though this issue can be circumvented using, for example, Firth's corrected estimator \citep{firth1993bias}.

>>>>

In this chapter we present the precise functional models and Bayesian methodologies explored in this work. The RKHS-based functional models under consideration \citep[see][]{berrendero2023functional, berrendero2019rkhs} are those obtained by substituting the functional parameter \(\beta\in L^2[0,1]\) for \(\alpha \in \Hcal(K)\), replacing also the scalar product \(\dotprod{X}{\beta}\) for \(\dotprod{X}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subspace \(\Hcal_0(K)\) defined in~\eqref{eq:h0}, i.e.:
\begin{equation}\label{eq:alpha_h0}
\alpha(\cdot) = \sum_{j=1}^p \beta_j K(t_j, \cdot), \text{ for some } p \in \N, \ \beta_j \in \R \text{ and } t_j \in [0,1].
\end{equation}
Moreover, as we said before, with a slight abuse of notation we will understand the expression \(\dotprod{x}{\alpha}_K\) as \(\Psi_x(\alpha)\), where \(x=X(\omega)\) and \(\Psi_x\) is Loève's isometry. Hence, taking into account that \(\Psi_X(K(t, \cdot)) = X(t)\) by definition (see~\eqref{eq:loeves-isometry}), when \(\alpha\) is as in~\eqref{eq:alpha_h0} we can write \(\dotprod{x}{\alpha}_K \equiv \sum_{j=1}^p \beta_j x(t_j)\).

In this way we get a simpler, finite-dimensional approximation of the functional RKHS model, which we argue reduces the overall complexity of the model while still capturing most of the relevant information. When it comes to parameter estimation, a direct optimization of some loss function would probably require a tailored algorithm that took into account the whole functional trajectories \(x(t)\) to select the appropriate times \(t_j\). Indeed, such an idea is explored in \citet{berrendero2023functional} for the logistic regression case, where the authors propose a ``greedy max-max'' method reminiscent of the EM algorithm that alternates between estimating the coefficients and the time instants through a maximum likelihood approach.

@@

In the case of functional logistic regression, we regard the response variable \(Y\in\{0, 1\}\) as a Bernoulli random variable given \(X=x \in L^2[0, 1]\), and as usual suppose that \(\log\left(p(x)/(1-p(x))\right)\) is linear in \(x\), where \(p(x)=\P(Y=1| X=x)\). Then, following the approach  suggested by \citet{berrendero2023functional}, a RKHS model might be given, in terms of the correspondence \(\dotprod{X}{\alpha}_K = \Psi_X(\alpha)\), by the equation
\begin{equation}\label{eq:rkhs-model-logistic}
  \P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\alpha}_K\}}, \quad \alpha_0 \in \R, \ \alpha \in \Hcal_{0,p}(K).
\end{equation}

@@

In much the same way as the linear regression model described above, this RKHS-based logistic regression model offers some advantages over the \(L^2\)-model~\eqref{eq:l2-logistic-model}. First and foremost, it has a more straightforward interpretation and allows for a workable Bayesian approach, as we will demonstrate below. Secondly, it can be shown that under mild conditions the general RKHS functional logistic model holds whenever the conditional distributions \(X | Y=0\) and \(X|Y=1\) are homoscedastic Gaussian processes, and in some cases it also entails the \(L^2\)-model \citep[see Theorem 1 in][]{berrendero2023functional}; this provides a sound theoretical motivation for the reduced model. Furthermore, a maximum likelihood approach for parameter estimation (although not considered here) is possible as well. Indeed, the use of a finite-dimensional approximation  mitigates the problem of non-existence of the MLE in the functional case. However, let us recall that even in finite-dimensional settings there are cases of quasi-complete separation in which the MLE does not exist \citep{albert1984existence}, though this issue can be circumvented using, for example, Firth's corrected estimator \citep{firth1993bias}.
